---
title: "Lab 03"
output: github_document
author: "Ken Field"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("NHANES")) install.packages("NHANES"); library(NHANES)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

## NHANES Dataset

Data from the US National Health and Nutrition Examination Study
https://www.cdc.gov/nchs/nhanes/index.htm

This is survey data collected by the US National Center for Health Statistics (NCHS) which has
conducted a series of health and nutrition surveys since the early 1960’s. Since 1999 approximately
5,000 individuals of all ages are interviewed in their homes every year and complete the health
examination component of the survey. The health examination is conducted in a mobile examination
center (MEC).

A description of the variables can be found in the package description.

```{r NHANES}
data("NHANES")
summary(NHANES)
help(NHANES)
```

## Model of Height Data

First look at the entire data set.

```{r Histogram}
ggplot(NHANES) +
  aes(x = Height) +
  geom_histogram(bins=100) +
  theme_cowplot()
```

Now lets just look at the data for children aged 2-17. 
Also drop any row that has NA for Height.

```{r Children Subset}
NHANES_child <- filter(NHANES, (Age >= 2 & Age <= 17))
NHANES_child <- drop_na(NHANES_child, Height)
ggplot(NHANES_child) +
  aes(x = Height) +
  geom_histogram(bins=100) +
  theme_cowplot()
```

## Model of the data

First, we will test to see if a simple model of using mean +- standard deviation
can predict the height of children. 

```{r Simple Model}
mean(NHANES_child$Height)
sd(NHANES_child$Height)
summary(NHANES_child$Height)
```

The error is large when using mean to predict the height. 
If this data had a normal distributation, this model would be better.

```{r}
error_mean <- NHANES_child$Height - mean(NHANES_child$Height)
ggplot(NULL, aes(error_mean)) +
  geom_histogram(bins = 100) + 
  xlim(-60, 60) +
  xlab("Error when predicting height with mean")
rmse_mean <- sqrt(mean(error_mean**2))
rmse_mean
```

More sophisticated models are necessary!

```{r Age}
p1 <- ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(position = "jitter",size=0.05) +
  scale_x_continuous(breaks = seq.int(0, 20, 2)) +
  ggtitle('A: original data') +
  theme_cowplot()

lmResultHeightOnly <- lm(Height ~ Age + 0, data=NHANES_child)
lmResultHeightOnly
summary(lmResultHeightOnly)
rmse_heightOnly <- sqrt(mean(lmResultHeightOnly$residuals**2))
rmse_heightOnly

p2 <-  ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(position = "jitter",size=0.05) +
  scale_x_continuous(breaks = seq.int(0, 20, 2)) + 
  annotate('segment',x=0,xend=max(NHANES_child$Age),
           y=0,yend=max(lmResultHeightOnly$fitted.values),
           color='blue',lwd=1) + 
  ggtitle('B: age') +
  theme_cowplot()

lmResultHeight <- lm(Height ~ Age, data=NHANES_child)
lmResultHeight
summary(lmResultHeight)
rmse_height <- sqrt(mean(lmResultHeight$residuals**2))
rmse_height

p3 <- ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(position = "jitter",size=0.05) +
  scale_x_continuous(breaks = seq.int(0, 20, 2)) + 
  geom_smooth(method='lm',se=FALSE) + 
  ggtitle('C: age + constant') +
  theme_cowplot()

plot_grid(p1,p2,p3,ncol=2)  ## Another great function of the cowplot package
```

This model is much better than using only the mean.

Can we make it even better?

```{r}
ggplot(NHANES_child, aes(x = Age, y = Height)) +
  geom_point(aes(colour = factor(Gender)), 
             position = "jitter", 
             alpha = 0.8,
             size=0.05) +
  geom_smooth(method='lm',aes(group = factor(Gender), 
                              colour = factor(Gender))) + 
  theme(legend.position = c(0.25,0.8)) + 
  ggtitle('D: age + constant + gender') +
  theme_cowplot()

lmResultGenderAge <- lm(Height ~ Gender + Age, data=NHANES_child)
lmResultGenderAge
summary(lmResultGenderAge)
```

## Test Assumptions


Acknowledgements
=========================

