---
title: "Group Project 1"
subtitle: "Biology 368/664 Bucknell University"
output: html_notebook
author: Derek Knight, Jessica Parrino, Margot Beck
date: 21 Feb 2022
---

```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("datasauRus")) install.packages("datasauRus"); library(datasauRus)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

This project will require you to develop a tutorial to teach Bucknell students how to use R for graphing and data analysis. 

## Target Audience

Bucknell 202 biology students interested in public health who are new to R and are looking to explore data analysis.  


## Grading

Each student will be expected to complete the following tasks to earn 85% of the points available for this assignment (21/25).

- Identify and obtain suitable dataset
- Use a Github repository and version control to collaborate on the project
- Spend 4-6 hours preparing, coding, and testing tutorial
  + Data exploration (Margot)
  + Data visualization (Derek)
  + Hypothesis testing
- Present tutorial in class
- Provide public archive suitable for sharing to students/faculty

Tutorials from previous classes can be viewed at our public github site: https://github.com/Bucknell-Biol364

Each group should use an *Acknowledgements* section to document the participation of each member and the collaboration within and between groups.

Additional credit will be awarded for providing assistance to other groups and for the development of a tutorial that goes beyond the minimal expectations listed above.

-----

## Introduction 
Welcome to coding in R, biology students! As scientists, you understand that data is everything when it comes to documenting the results of your biology labs, supporting or rejecting a hypothesis in an experiment, or understanding trends in public health (like perhaps transmission trends in a worldwide pandemic).Data doesn't speak for itself though, it requires a little bit of help on our part. What good does a spreadsheet of numbers, characters, or observations do on its own if you don't perform some analyses to make sense of it? That is where programs like R come in handy to scientists (or anyone working with datasets, for that matter) as important tools for data exploration, modeling, and analysis. By following along with this R tutorial, you will begin to see how useful coding can be when it comes to working with biological data.  

## About the Dataset
The data you will be using for this exercise is the NHANES 2009-2012 dataset.  It is survey data from the US National Center for Health Statistics that interviews approximately 5,000 individuals. The interviewees also complete a health examination as part of the survey.  75 variables are included for the 2009-2010 and 2011-2012 sample years.  

## Download Packages and Obtain the Dataset 
First, you will need to download all required packages and download the NHANES dataset. Packages are essentially collections of functions and data sets. Run the first code chunk below by clicking the green arrow in the top right hand corner to install the packages. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("NHANES")) install.packages("NHANES"); library(NHANES)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

Next, run the below chunk of code line by line.  You can do this by hitting "Ctrl+Enter" on each line. You can also click the green arrow in the upper right hand corner of grey box. The first line will load the dataset "NHANES" into your Environment tab (upper right hand window of R-studio). The second line will open up an info page with more information about the NHANES dataset in the Help tab of the lower right hand window of R-studio.

```{r NHANES}
data("NHANES")
help(NHANES)
```


## Explore the Dataset
Now that you have the relevant packages installed and the dataset downloaded, it's time to get an idea of what the data are made up of. Run the below code chunk line by line to explore the data. The first function, "summary", will provide you with a summary of the dataset which includes the minimum, median, mean, maximum, 1st quartile, 3rd quartile, and NA's for each variable. In the next two lines, the function "head" will show you the first 6 rows of each variable while "tail" will show you the last 6. In the fourth line of code, the function "dim" will display the dimensions of the data table in the form of 'row, column'. This code chunk should help to give you an idea of what variables make up the data, the size of the dataset, and what the range of observations are for each variable.  

```{r Data Exploration}
summary(NHANES)
head(NHANES)
tail(NHANES)
dim(NHANES)
```


## Sample Dataset

One of the possible datasets to use for the tutorial can be found in the datasauRus package.

```{r}
datasaurus_dozen %>% 
  group_by(dataset) %>% 
  summarize(
      mean_x    = mean(x),
      mean_y    = mean(y),
      std_dev_x = sd(x),
      std_dev_y = sd(y),
      corr_x_y  = cor(x, y)
      )
```

Boxplots of either the x or the y value show that there is something strange going on.

```{r}
ggplot(datasaurus_dozen, aes(x = x, colour = dataset))+
    geom_boxplot()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

```{r}
ggplot(datasaurus_dozen, aes(x = y, colour = dataset))+
    geom_boxplot()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

But you have to visualize all of the data with a scatter plot to really see the patterns.

```{r fig.height=12, fig.width=9}
  ggplot(datasaurus_dozen, aes(x = x, y = y, colour = dataset))+
    geom_point()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

And did you notice the code in the {r} codechunk header that controlled the size of the output in the Rmd? Pretty neat trick!

And here are two versions of the data that you could use in your data visualization tutorial. 
To use them you would probably want to change the names of the datasets and also make x and y more meaningful. 
Then save them as a csv or tsv to be imported later for your tutorial. 

```{r}
datasaurus_long <- datasaurus_dozen
datasaurus_wide <- datasaurus_dozen_wide
head(datasaurus_long)
head(datasaurus_wide)
```

# Acknowledgements
The information about the NHANES dataset was found on the 'help' page for NHANES in R.

Margot Beck wrote the introduction and data exploration sections.  

Code used for exploring the data was inspired by a blog post on R-bloggers (https://www.r-bloggers.com/2018/11/explore-your-dataset-in-r/).


DatasauRus package and description below: Stephanie Locke https://github.com/jumpingrivers/datasauRus

The datasauRus package wraps the awesome Datasaurus Dozen dataset, which contains 13 sets of x-y data. Each sub-dataset has five statistics that are (almost) the same in each case. (These are the mean of x, mean of y, standard deviation of x, standard deviation of y, and Pearson correlation between x and y). However, scatter plots reveal that each sub-dataset looks very different. The dataset is intended to be used to teach students that it is important to plot their own datasets, rather than relying only on statistics.

The Datasaurus was created by Alberto Cairo in this great [blog post](http://www.thefunctionalart.com/2016/08/download-datasaurus-never-trust-summary.html).

He's been subsequently made even more famous in the paper [Same Stats, Different Graphs: Generating Datasets with Varied Appearance and Identical Statistics through Simulated Annealing](https://www.autodeskresearch.com/publications/samestats) by Justin Matejka and George Fitzmaurice.

In the paper, Justin and George simulate a variety of datasets that the same summary statistics to the Datasaurus but have very different distributions. 

This package also makes these datasets available for use as an advanced [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet), available in R as `anscombe`.
